A Theory for Database Transactions with no Database

### 2019-04-03 2018 Transactions/articleTransactions.tex
WIP

\section{Introduction}
\label{sct:Introduction}
	A distributed information system (DIS) serves its users with things such as
	registering data, guarding the consistency, protecting the data,
	and providing that data accurately and meaningfully for its intended purpose.
	Users want predictable behaviour when they are simultaneously changing a shared state (e.g.\ flight reservation).
	For that purpose a transaction mechanism preserves system-wide invariants.
	Many transaction mechanisms are known that presume a shared database [refs].
	However, shared databases are not always an option in distributed cloud architectures.
	Globally distributed applications such as Netflix and LinkedIn cannot afford to rely on a shared database for obvious reasons of performance and reliability.
	Many cloud applications use distributed storage models instead.
	Examples are event stores (e.g.\ Kafka), triple stores (e.g.\ semantic web), graph databases (e.g. Neo4J) and others.
	These so-called NoSQL databases often lack functionality for transactions.
	For the purpose of doing transactions in a non-transactional situation, several architectural patterns exist.
	Some of these patterns simply circumvent the problem.
	Others will isolate transactionality into a smaller sub-problem.
	Still others introduce locking mechanisms to do the job.
	Yet, in the world of event-driven distributed systems, a dire need is felt for a theory that brings what transactions do for shared databases.

	This paper presents such a theory.
	It describes transactions without the assumption of a shared database.
	In this theory we express the characteristic properties of transactions: Atomicity, Consistency, Isolation, and Durability.
	{\em Atomicity} means that a sequence of primitive actions can be executed as though it were a single action.
	{\em Consistency} means that integrity rules are warranted by the database management system.
	{\em Isolation} means that actions can be done without any knowledge of what other users are doing simultaneously.
	{\em Durability} means that the contents of a database cannot change unintendedly.

@@

\section{History}
\label{sct:History}
	Distributed information systems are dynamic systems,
	so time is the first aspect we focus on.
	Being distributed means that we cannot assume that the temporal order of events is known in all cases.
	For this reason we draw on a theory of concurrency, which is generally applicable,
	well-established, and founded in a solid mathematical structure~\cite{Foster14}.
	In this paper we simplify the notion of time to precedence of events,
	allowing us to study the evolution over time by events that happen.
	We need the notion of {\em event}\footnote{Cambridge Dictionary defines ``event'' as: anything that happens} to describe things that happen in a DIS.
	In this work we use variables $e$, $e_1$, and $e_2$ to represent events.
	Let $\mathbb E$ represent the set of all conceivable events.
	We also introduce the notion of {\em history structure}
	to represent all events that have happend up to some point in time and the precedences between those events.
	In this paper, variables $h$, $h'$, $h_1$, and $h_2$ represent history structures.
	Let $\mathbb H$ be the set of all conceivable history structures.
	We describe the history structure of a DIS as a partial order of events

>>>>

\section{Introduction}
\label{sct:Introduction}
	A distributed information system (DIS) serves its users with things such as
	registering data, guarding the consistency, protecting the data,
	and providing that data accurately and meaningfully for its intended purpose.
	Users want predictable behaviour when they are simultaneously changing a shared state (e.g.\ flight reservation).
	For that purpose a transaction mechanism preserves system-wide invariants.
	Many transaction mechanisms are known that presume a shared database [refs].
	However, shared databases are not always an option in distributed cloud architectures.
	Globally distributed applications such as Netflix and LinkedIn cannot afford to rely on a shared database for obvious reasons of performance and reliability.
	Many cloud applications use distributed storage models instead.
	Examples are event stores (e.g.\ Kafka), triple stores (e.g.\ semantic web), graph databases (e.g. Neo4J) and others.
	These so-called NoSQL databases often lack functionality for transactions,
	leaving the problems for the user to solve.
	As a result, many architectural patterns exist for the purpose of doing transactions in a distributed setting.
	Some of these patterns simply circumvent the problem.
	Others will isolate transactionality into a smaller sub-problem.
	Still others introduce locking mechanisms to do the job.
	In paper we will show that the ACID-properties of transactions follow quite naturally from the theory of time.
	In informal terms: the merge of Git corresponds to the transactional commit in databases.
	This result contributes to the body of foundational theory of information systems.

	This paper describes transactions without the assumption of a shared database.
	In we express the characteristic properties of transactions: Atomicity, Consistency, Isolation, and Durability.
	{\em Atomicity} means that a sequence of primitive actions can be executed as though it were a single action.
	{\em Consistency} means that integrity rules are warranted by the database management system.
	{\em Isolation} means that actions can be done without any knowledge of what other users are doing simultaneously.
	{\em Durability} means that the contents of a database cannot change unintendedly.
	For this purpose we build on a well-known theory of time~\cite{DBLP:journals/ipl/FosterCWZ18},
	which is generally applicable,
	well-established, and founded in a solid mathematical structure~\cite{Foster14}.

@@

\section{History}
\label{sct:History}
	Distributed information systems are dynamic systems,
	so time is the first aspect we focus on.
	Being distributed means that we cannot assume that the temporal order of events is known in all cases.
	In this paper we simplify the notion of time to precedence of events,
	allowing us to study the evolution over time by events that happen.
	We need the notion of {\em event}\footnote{Cambridge Dictionary defines ``event'' as: anything that happens} to describe things that happen in a DIS.
	In this work we use variables $e$, $e_1$, and $e_2$ to represent events.
	Let $\mathbb E$ represent the set of all conceivable events.
	We also introduce the notion of {\em history structure}
	to represent all events that have happend up to some point in time and the precedences between those events.
	In this paper, variables $h$, $h'$, $h_1$, and $h_2$ represent history structures.
	Let $\mathbb H$ be the set of all conceivable history structures.
	We describe the history structure of a DIS as a partial order of events
